---
title: "biolabs_project"
output:
  html_document:
    df_print: paged
date: "2022-06-23"
---

Creating MES.5

```{r tyding data -> MES.5}
library(tidyverse)
library(tidyr)

#Remove _Methyl_100ng_ from column names
MES -> MES.5
colnames(MES.5) <- gsub("_Methyl_100ng_", "_", colnames(MES.5))

#rename and separate #KEY into gene_id and type 
MES.5 %>% 
  separate(`#KEY`, c("gene_id", "type"), 
           sep = "\\|",
           remove = T, 
           convert = T) -> MES.5

#change _ in gene_id to -
MES.5$gene_id <- gsub('_', '-', MES.5$gene_id)

#separate gene_id into gene_id and suffix
##some gene_id have two - or no -
MES.5 %>%
  rename(isoform = gene_id)|>
  separate(isoform, c("gene_id", "suffix"),
           sep = "-(?!.*-)",
           remove = F, 
           convert = T,
           fill = "right") -> MES.5

#Remove all rows with pseudocounts of 0 over all days, experiments and replicates 
MES.5[rowSums(MES.5[, 5:48])>0,] -> MES.5

#Remove numerical NAs
which(is.na(MES.5), arr.ind=TRUE)
MES.5[-59652,] -> MES.5

```

```{r MES.5.csv}

MES.5 %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES.5.csv")

```

Creating MES1

```{r MES.5 -> MES1}

#Pivot longer to assign case_id to each transcript and create row of pseudocounts
MES.5 %>% 
  pivot_longer(cols = c(5:48), names_to = "case_id", values_to = "pseudocounts") -> MES1

#Remove KK_MES_ from the front and -2 from the end of the string 
#Separated case_id into 3 columns (experiment, day and rep)
MES1 %>% 
  mutate(case_id = str_sub(case_id, start = 8, end = -3)) %>% 
  separate(case_id, c("exp", "day", "rep"), 
           sep = "_", 
           remove = T, 
           convert = T,
           fill = "left") -> MES1

#Replace NAs with Exp1, remove the exp and remove the D infront of day number
MES1 <-
  replace_na(MES1, list(exp = "Exp1"))
MES1 <- 
  mutate(MES1, exp = str_sub(exp, start = 4)) 
MES1 <- 
  mutate(MES1, day = str_sub(day, start = 2))

#Change exp and day from characters to numeric 
MES1 <- 
  mutate(MES1, across(.cols = 5:6, .fns = as.numeric))

```

```{r MES1.csv}

MES1 %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES1.csv")

```

Creating MES2

```{r MES1 -> MES2}

#pivot wider to have days 0-10 as variables 
MES1 %>% 
  pivot_wider(names_from = day, values_from = pseudocounts) -> MES2

#order columns in ascending order of day
MES2 <- MES2[c(1, 2, 3, 4, 5, 6, 17, 16, 15, 14, 13, 12, 11, 10, 9, 8, 7)]

#rename day variables to add the word day  (probably and easier way)
MES2 %>% 
  rename(day_0 = '0',
         day_1 = '1',
         day_2 = '2',
         day_3 = '3',
         day_4 = '4',
         day_5 = '5',
         day_6 = '6',
         day_7 = '7',
         day_8 = '8',
         day_9 = '9',
         day_10 = '10') -> MES2

#arrange replicates and experiments in ascending order, grouped by gene_id and suffix
MES2%>% 
  group_by(gene_id, suffix) %>% 
  arrange(exp, rep, .by_group = T) -> MES2

#mutate new total sum column of the total psuedocounts for each replicate across all days
MES2 %>% 
  rowwise() %>%
  mutate(total = sum(c_across(day_0:day_10), na.rm = T)) -> MES2

```

```{r MES2.csv}

MES2 %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES2.csv")

```

```{r Find primary key}

MES.5 %>% 
  count(gene_id, suffix) %>% 
  arrange(desc(n))

MES1 %>% 
  count(gene_id, suffix, exp, day, rep) %>% 
  arrange(desc(n))

MES2 %>% 
  count(gene_id, suffix, exp, rep) %>% 
  arrange(desc(n))

```

Looking at top types of RNA

```{r Find most important types}
#viewing number of each type and the respective pseudocounts arranged by sum
temp <- MES1|>
  group_by(type)|>
  summarise(n = n(),
            sum = sum(pseudocounts))|>
  arrange(desc(sum))
view(temp)
```

```{r Graph of top 10 most prevalent types with standard deviation bars}
#Creating a graph based on the average pseudocount for each of the top 10 occurring types, over each of the 11 days
MES1sum <- MES1|>
  group_by(day,rep,exp,type)|>
  summarize(pseudocounts=sum(pseudocounts))|>
  filter(type == "rRNA" | type=="protein_coding" | type == "misc_RNA" | type == "lincRNA" | type == "retained_intron" | type == "processed_transcript"
         | type == "nonsense_mediated_decay" | type == "processed_pseudogene" | type == "ribozyme" | type == "TEC")|>
  group_by(rep,day,exp)|>
  summarize(across(average=mean(pseudocounts)))|>
  group_by(type,day)|>
  summarize(avg=mean(pseudocounts),sd=sd(pseudocounts))

#Reordering the types based on final position on graph

MES1sum$type <- factor(MES1sum$type, levels=c("protein_coding", "rRNA", "misc_RNA","lincRNA", "retained_intron", "nonsense_mediated_decay", "processed_pseudogene","processed_transcript", "ribozyme", "TEC"))

#plotting the data
ggplot(MES1sum,mapping=aes(x=day,y=log(avg), color=type))+
  geom_line()+
  geom_point()+
  geom_errorbar(aes(x=day,y=log(avg),ymin=log(avg-sd), ymax=log(avg+sd),width=.2))+
  scale_x_continuous(breaks = 0:10)+
  labs(title="The log Average Pseudocount of Transcript Type Over 11 Days",
       subtitle = "The log average of the top 10 most prevalent RNA types based on pseudocount sum, over the 11 days",
       x="Day",y="log(avg)", col="RNA Type")

# ggsave("/Users/danielle.dyer/Desktop/avg_days.png", dpi = 600)
```

```{r another plot of same graph only with standard error bars}

# Creating the plotting tibble

## Finding ten most "abundant" transcript types
MES1 |>
  group_by(type) |>
  summarize(total = sum(pseudocounts)) |>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(type) ->
  top_ten

## Day 11 ranking
MES1 |>
  filter(day == 10) |>
  group_by(type) |>
  summarize(total = sum(pseudocounts)) |>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(type) ->
  top_ten
  

## Filtering by top ten, summarizing and plotting
MES1 |>
  filter(type %in% top_ten) |>
  group_by(type, exp, rep, day) |>
  summarize(total = sum(pseudocounts)) |>
  mutate(type = factor(type)) |>
  ggplot(mapping = aes(x = day, y = total, color = type)) +
  stat_summary(geom = "line", fun = mean) +
  stat_summary(fun.data = mean_se) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_brewer(palette = "Set3", name = "RNA Type") +
  labs(title="The log Average Pseudocount of Transcript Type Over 11 Days",
       subtitle = "The top 10 most prevalent RNA types based on pseudocount sum",
       x = "Day", y = "Pseudocount") +
  coord_trans(y = "log10") +
  theme_minimal()

```

Should the data be normalized

```{r normalizing MES.5}
library(dplyr)

#create function
normalize <- function(x){
  x/sum(x, na.rm = T)
}

MES.5|>
  mutate(across(5:48, .fns = normalize)) ->
  MES.5_norm


#testing 
summary(colSums(MES.5[5:48]))

sd(colSums(MES2[7:17], na.rm = T))



```

```{r MES.5_norm.csv}

MES.5_norm %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES.5_norm.csv")

```

```{r Normalize MES1}
MES.5_norm %>% 
  pivot_longer(cols = c(5:48), names_to = "case_id", values_to = "pseudocounts") -> MES1_norm

#Remove KK_MES_ from the front and -2 from the end of the string 
#Separated case_id into 3 columns (experiment, day and rep)
MES1_norm %>% 
  mutate(case_id = str_sub(case_id, start = 8, end = -3)) %>% 
  separate(case_id, c("exp", "day", "rep"), 
           sep = "_", 
           remove = T, 
           convert = T,
           fill = "left") -> MES1_norm

#Replace NAs with Exp1, remove the exp and remove the D infront of day number
MES1_norm <-
  replace_na(MES1_norm, list(exp = "Exp1"))
MES1_norm <- 
  mutate(MES1_norm, exp = str_sub(exp, start = 4)) 
MES1_norm <- 
  mutate(MES1_norm, day = str_sub(day, start = 2))

#Change exp and day from characters to numeric 
MES1_norm <- 
  mutate(MES1_norm, across(.cols = 5:6, .fns = as.numeric))


```

```{r MES1_norm.csv}

MES1_norm %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES1_norm.csv")

```

Try the previous graphs with the normalized data

```{r Create same graph with MES1_norm}
# Creating the plotting tibble

## Finding ten most "abundant" transcript types
MES1_norm |>
  group_by(type) |>
  summarize(total = sum(pseudocounts)) |>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(type) ->
  top_ten

## Day 11 ranking
MES1_norm |>
  filter(day == 10) |>
  group_by(type) |>
  summarize(total = sum(pseudocounts)) |>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(type) ->
  top_ten
  

## Filtering by top ten, summarizing and plotting
MES1_norm |>
  filter(type %in% top_ten) |>
  group_by(type, exp, rep, day) |>
  summarize(total = sum(pseudocounts)) |>
  mutate(type = factor(type)) |>
  ggplot(mapping = aes(x = day, y = total, color = type)) +
  stat_summary(geom = "line", fun = mean) +
  stat_summary(fun.data = mean_se) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_brewer(palette = "Set3", name = "RNA Type") +
  labs(x = "Day", y = "Pseudocount") +
  coord_trans(y = "log10") +
  theme_minimal()+
  theme_dark()+
  theme(panel.grid.major = element_line(color = "white"),
        panel.grid.minor = element_line(color = "white"))
```

Back to normalizing the last data set created

```{r Normalizing MES2}

#pivot wider to have days 0-10 as variables 
MES1_norm %>% 
  pivot_wider(names_from = day, values_from = pseudocounts) -> MES2_norm

#order columns in ascending order of day
MES2_norm <- MES2_norm[c(1, 2, 3, 4, 5, 6, 17, 16, 15, 14, 13, 12, 11, 10, 9, 8, 7)]

#rename day variables to add the word day  (probably and easier way)
MES2_norm %>% 
  rename(day_0 = '0',
         day_1 = '1',
         day_2 = '2',
         day_3 = '3',
         day_4 = '4',
         day_5 = '5',
         day_6 = '6',
         day_7 = '7',
         day_8 = '8',
         day_9 = '9',
         day_10 = '10') -> MES2_norm

#arrange replicates and experiments in ascending order, grouped by gene_id and suffix
MES2_norm %>% 
  group_by(gene_id, suffix) %>% 
  arrange(exp, rep, .by_group = T) -> MES2_norm

#mutate new total sum column of the total psuedocounts for each replicate across all days
MES2_norm %>% 
  rowwise() %>%
  mutate(total = sum(c_across(day_0:day_10), na.rm = T)) -> MES2_norm

```

```{r MES2_norm.csv}

MES2_norm %>% 
  write_csv("/Users/danielle.dyer/Documents/DS Practicum/BioLabs Project/CSV/MES2_norm.csv")

```

```{r Graph of top 10 transcipts within experiment 1 and replicate 1 by psuedocount across 11 days (un-normalized data)}
# identify and create list of top 10 transcripts with highest total pseudocounts from experiment 1 replicate 1 
MES1|>
  filter(exp == "1", rep == "1")|>
  group_by(isoform)|>
  summarize(total = sum(pseudocounts))|>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(isoform) ->
  top_ten_transcript

#plot this top 10 subset faciting by each experiment and replicate
MES1 |>
  filter(isoform %in% top_ten_transcript) |>
  ggplot(mapping = aes(x = day, y = pseudocounts, color = isoform)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(exp), cols = vars(rep)) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_brewer(palette = "Set3", name = "Isoform") +
  labs(x = "Day", y = "Pseudocount") +
  coord_trans(y = "log10") +
  theme_minimal()

# ggsave("/Users/danielle.dyer/Desktop/top_ten_transcripts.png", dpi = 600)

```

```{r same as previous graph (normalized data)}

MES1_norm|>
  filter(exp == "1", rep == "1")|>
  group_by(isoform)|>
  summarize(total = sum(pseudocounts))|>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(isoform) ->
  top_ten_transcript

MES1_norm |>
  filter(isoform %in% top_ten_transcript) |>
  ggplot(mapping = aes(x = day, y = pseudocounts, color = isoform)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(exp), cols = vars(rep)) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_brewer(palette = "Set3", name = "Isoform") +
  labs(x = "Day", y = "Pseudocount") +
  # coord_trans(y = "log10") +
  theme_minimal()

# ggsave("/Users/danielle.dyer/Desktop/top_ten_transcripts_norm.png", dpi = 600)

```



Looking at percent changes from day to day within the top_ten_transcript subset of the data
```{r Percent change from day to day}

#Creating a column at the end of the data table for percent change for top ten transcripts based on pseudocounts
MES1_pct <- MES1|>
  filter(isoform%in%top_ten_transcript)|>
  group_by(isoform, exp, rep)|> #This is so the percent changes are based on previous values within the same transcript, experiment number and repetition number
  arrange(day)|>
  mutate(pct_change = ((pseudocounts/lag(pseudocounts))-1)*100)

#Omitting the transcript IDs "Eef1a1_201" and "Col4a1_208" due to strangely high changes in pseudocounts (48,201,735% was highest)
#comment this code out if wanting to see the these transcripts
MES1_pct <- MES1_pct|>
  filter(isoform != "Col4a1-208" & isoform != "Eef1a1-201")

#Graphing the percent change in pseudocount, facet wrapped so that rows are experiment number and columns are repetitions
MES1_pct|>
  ggplot(mapping = aes(x=day, y = pct_change, color=isoform))+
  geom_line()+
  facet_grid(rows = vars(exp), cols = vars(rep)) +
  scale_x_continuous(breaks = 0:10) +
  scale_color_brewer(palette = "Set3", name = "Isoform") +
  labs(x = "Day", y = "Percent Change") +
  theme_minimal()
```


Model of top 10 isoforms with the greatest absolute difference (original data)
```{r create mes_filter of top 10 highest absolute differences between the first and second half of the days}
MES1|>
  mutate(group = ifelse(day < 5, "first", "second")) |>
  group_by(isoform, group)|>
  summarize(avg = mean(pseudocounts)) -> mes_diff

mes_diff|>
  pivot_wider(names_from = group, values_from = avg)|>
  mutate(diff = second - first) -> mes_diff

mes_diff |>
  group_by(isoform)|>
  summarize(diff = diff)|>
  arrange(desc(abs(diff))) |>
  slice(1:10) -> diff_filter

mes_diff |>
  group_by(isoform)|>
  summarize(diff = diff)|>
  arrange(desc(abs(diff))) |>
  slice(1:10) |>
  pull(isoform) -> top10_diff



```

```{r model/graph diff_filter}
MES1|>
  semi_join(diff_filter) -> diff_filter1 

MES1|>
  semi_join(diff_filter) |>
  filter(isoform != "Col4a1-208") -> 
  diff_filter2 
# isoform != "Col4a1-201",

#model for filter 1 with colagen RNA
mod_1 <- lm(log(pseudocounts) ~ day * isoform, data = diff_filter1)
# summary(mod_1)

# plot(mod_1)

diff_filter1 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")


#model for filter 2 without colagen RNA
mod_2 <- lm(log(pseudocounts) ~ day * isoform, data = diff_filter2)
# summary(mod_1)

# plot(mod_2)

diff_filter2 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")

```

```{r self fitted models}

#own model plotted step by step
# fitted.values(mod_2)
mod_2_plot_df <- diff_filter2
mod_2_plot_df$preds <- fitted.values(mod_2)
mod_2_plot_df$pseudocounts <- log(mod_2_plot_df$pseudocounts)

mod_2_ci <- as_tibble(predict(object = mod_2,
                              interval = "confidence")[, 2:3])

mod_2_plot_df |>
  add_column(mod_2_ci) ->
  mod_2_plot_df

mod_2_plot_df$isoform <- factor(mod_2_plot_df$isoform, levels=c("unit-45s-5ets", "unit-45s-its2", "Col4a1-201", "Gm26917-202", "unit-45s-its1", "Col4a2-201", "Lama1-201", "Lamc1-201", "Rpph1-201"))

mod_2_plot_df |>
  ggplot(aes(x = day, y = pseudocounts)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = isoform), show.legend = F, alpha = 0.2) +
  # geom_point(aes(color = isoform)) +
  geom_line(aes(y = preds, color = isoform))+
  scale_x_continuous(breaks = 0:10) +
  labs(title="Linear models of the top 9 transcripts with the highest absolute difference",
       x = "Day", y = "Pseudocount", col = "Isoform")
  
  


#own fitted model with all 10 from mod 1
# fitted.values(mod_1)
mod_1_plot_df <- diff_filter1
mod_1_plot_df$preds <- fitted.values(mod_1)
mod_1_plot_df$pseudocounts <- log(mod_1_plot_df$pseudocounts)

mod_1_ci <- as_tibble(predict(object = mod_1,
                              interval = "confidence")[, 2:3])

mod_1_plot_df |>
  add_column(mod_1_ci) ->
  mod_1_plot_df

mod_1_plot_df$isoform <- factor(mod_1_plot_df$isoform, levels=c("unit-45s-5ets", "unit-45s-its2", "Col4a1-201", "Gm26917-202", "unit-45s-its1", "Col4a2-201", "Lama1-201", "Lamc1-201", "Rpph1-201", "Col4a1-208"))

mod_1_plot_df |>
  ggplot(aes(x = day, y = pseudocounts)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = isoform), show.legend = F, alpha = 0.3) +
  geom_point(aes(color = isoform), position = "jitter") +
  geom_line(aes(y = preds, color = isoform))+
  scale_x_continuous(breaks = 0:10) +
  labs(title="Linear models of the top 10 transcripts with the highest absolute difference",
       x = "Day", y = "Pseudocount", col = "Isoform")
  
  


```

```{r pairwise comparison of trends}
# library(emmeans)

emtrends(object = mod_2,
         specs = pairwise ~ isoform,
         var = "day")
```

```{r making readable table from results}
library(gt)

#Absolute difference table using mod 2 (diff_filter2)
try2 <- emtrends(object = mod_2,
                 specs = pairwise ~ isoform,
                 var = "day")
contrast2 <- as.data.frame(try2$contrasts)

contrast2|>
  mutate(group = str_split(contrast, pattern = " - "),
         group = map(group, ~ .x[[1]][1]),
         group = str_sub(group, start = 2, end = -2))|>
  arrange(desc(estimate))|>
  gt(groupname_col = "group")|>
  fmt_number(columns = c("estimate", "SE", "t.ratio", "p.value"),
             decimals = 3)|>
  tab_options(column_labels.background.color = "#9999FF")|>
  tab_options(row_group.background.color = "#99CCFF")|>
  tab_options(row_group.font.weight = "bold")|>
  tab_header(
    title = "Absolute Difference"
  )|>
  tab_options(
    heading.title.font.size = "30px",
    heading.title.font.weight = "bold")|>
  data_color(
    columns = estimate,
    colors = "YlGnBu")|>
  tab_style(
    style = cell_fill(color = "red"),
    locations = cells_body(
      columns = p.value,
      rows = p.value >= 0.05)
  )|>
  tab_style(
    style = cell_fill(color = "green"),
    locations = cells_body(
      columns = p.value,
      rows = p.value <= 0.05))

```

Model of top ten isoforms with the highest total pseudocounts
```{r find top 10 transcripts with highest pseudocounts}

MES1|>
  group_by(isoform)|>
  summarize(total = sum(pseudocounts))|>
  arrange(desc(total)) |>
  slice(1:10) |>
  pull(isoform) ->
  top10_isoform

MES1|>
  group_by(isoform)|>
  summarize(total = sum(pseudocounts))|>
  arrange(desc(total)) |>
  slice(1:10) -> sum_filter


```

```{r model/graph sum_filter}
MES1|>
  semi_join(sum_filter) -> sum_filter1 

#no log transformation
mod_3.5 <- lm(pseudocounts ~ day * isoform, data = sum_filter1)
# summary(mod_3.5)
# plot(mod_3.5)

mod_3 <- lm(log(pseudocounts) ~ day * isoform, data = sum_filter1)
# summary(mod_3)
# plot(mod_3)

#linear model plot from ggplot
sum_filter1 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")

```

```{r self fitted models}
#own model plotted step by step
# fitted.values(mod_3)
mod_3_plot_df <- sum_filter1
mod_3_plot_df$preds <- fitted.values(mod_3)
mod_3_plot_df$pseudocounts <- log(mod_3_plot_df$pseudocounts)


mod_3_ci <- as_tibble(predict(object = mod_3,
                              interval = "confidence")[, 2:3])

mod_3_plot_df |>
  add_column(mod_3_ci) ->
  mod_3_plot_df

mod_3_plot_df$isoform <- factor(mod_3_plot_df$isoform, levels=c("unit-45s-5ets", "unit-45s-its2", "Rn7s1-201", "Gm26917-202", "unit-45s-its1", "Rn28S", "Rn7sk-201", "Gm42418-201", "Eef1a1-201", "Rpph1-201"))

mod_3_plot_df |>
  ggplot(aes(x = day, y = pseudocounts)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = isoform), show.legend = F, alpha = 0.5) +
  # geom_point(aes(color = isoform)) +
  geom_line(aes(y = preds, color = isoform))+
  scale_x_continuous(breaks = 0:10) +
  labs(title="Linear models of the top 10 most abundent transcripts",
       x = "Day", y = "Pseudocount", col = "Isoform")
  
```

```{r pairwise comparison of trends}
# library(emmeans)

emtrends3 <- emtrends(object = mod_3,
         specs = pairwise ~ isoform,
         var = "day")
```

```{r making readable table from results}
#Difference between slopes table using mod 3 (sum_filter1)
try3 <- emtrends(object = mod_3,
                  specs = pairwise ~ isoform,
                  var = "day")
contrast3 <- as.data.frame(try3$contrasts)
slopes <- as.data.frame(try3$emtrends)

contrast3|>
  mutate(group = str_split(contrast, pattern = " - "),
         group = map(group, ~ .x[[1]][1]),
         group = str_sub(group, start = 2, end = -2))|>
  arrange(desc(estimate))|>
  gt(groupname_col = "group")|>
  fmt_number(columns = c("estimate", "SE", "t.ratio", "p.value"),
             decimals = 3)|>
  tab_options(column_labels.background.color = "#9999FF")|>
  tab_options(row_group.background.color = "#99CCFF")|>
  tab_options(row_group.font.weight = "bold")|>
  tab_header(
    title = "Highest Pseudocounts",
    subtitle = "Estimates of how different the slopes of the transcripts are from each other"
  )|>
  tab_options(
    heading.title.font.size = "30px",
    heading.title.font.weight = "bold")|>
  data_color(
    columns = estimate,
    colors = "YlGnBu")|>
  tab_style(
    style = cell_fill(color = "red"),
    locations = cells_body(
      columns = p.value,
      rows = p.value >= 0.05)
  )|>
  tab_style(
    style = cell_fill(color = "green"),
    locations = cells_body(
      columns = p.value,
      rows = p.value <= 0.05))

```


Filter for a top 10: by maximum difference between sums between days (combining all reps), take the top 10 max differences
```{r find top 10 isoforms with highest daily difference counts}
MES1_norm %>%
  select(isoform, type, exp, day, pseudocounts) %>%
  group_by(day, isoform) %>%
  mutate(sum = sum(pseudocounts), .keep = "unused") %>%
  distinct() -> daily_diff

daily_diff %>%
  group_by(isoform, exp) %>%
  mutate(abs_daily_diff = abs(lead(sum) - sum)) -> daily_diff

daily_diff |>
  group_by(isoform)|>
  filter(abs_daily_diff != "NA") |>
  summarize(total = sum(abs_daily_diff))|>
  arrange(desc(total)) |>
  slice(1:10) -> spike_filter

daily_diff |>
  group_by(isoform)|>
  filter(abs_daily_diff != "NA") |>
  summarize(total = sum(abs_daily_diff))|>
  arrange(desc(total)) |>
  slice(1:10)|>
  pull(isoform) -> top10_spike
```

```{r model/graph spike_filter}
MES1_norm|>
  semi_join(spike_filter) -> spike_filter1

MES1_norm|>
  semi_join(spike_filter) |>
  filter(isoform != "Col4a1-201") -> 
  spike_filter2 

MES1_norm|>
  semi_join(spike_filter) |>
  filter(isoform != "Col4a1-201", isoform != "Col4a2-201") -> 
  spike_filter3

#model for filter 1 with colagen RNA
mod_4 <- lm(log(pseudocounts) ~ day * isoform, data = spike_filter1)
summary(mod_4)

plot(mod_4)

spike_filter1 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")


#model for filter 2 without 1 colagen RNA
mod_5 <- lm(log(pseudocounts) ~ day * isoform, data = spike_filter2)
summary(mod_5)

plot(mod_5)

spike_filter2 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")


#model for filter 3 with both colagen RNA
mod_6 <- lm(log(pseudocounts) ~ day * isoform, data = spike_filter3)
summary(mod_6)

plot(mod_6)

spike_filter3 |>
  ggplot(aes(x = day, y = log(pseudocounts), color = isoform)) +
  geom_point(aes(shape = factor(exp)), position = "jitter") +
  geom_smooth(method = "lm", se = F) +
  scale_color_brewer(palette = "Set3")


```


Look for differences between experiments and replicates 
```{r graphs for difference just between experiment}

#reminder: sum_filter1 = top 10 transcripts with highest pseudocounts in MES1 format
#graph comparing experiments with mean and standard error bars
sum_filter1 |>
  ggplot(aes(x = factor(exp), y = log(pseudocounts), color = factor(exp))) +
  # geom_point(color = "lightgray")+
  stat_summary(show.legend = F)+
  labs(title = "Difference between experiments", x = "Experiment")

#graph comparing experiments with mean and all summary points
sum_filter1 |>
  ggplot(aes(x = factor(exp), y = log(pseudocounts), color = factor(exp))) +
  geom_point(color = "lightgray")+
  stat_summary(show.legend = F)+
  labs(title = "Difference between experiments", x = "Experiment",)


##note that these graphs are on different scales!

```
Using sum_filter1 subset to try different models 
```{r model with isoform and experiment interaction }
#model
mod_7 <- lm(log(pseudocounts) ~ factor(exp) * isoform, data = sum_filter1)
summary(mod_7)

##graph
sum_filter1 |>
  ggplot(aes(x = isoform, y = log(pseudocounts), color = factor(exp))) +
  # geom_point(color = "lightgray")+
  stat_summary()

emmean7 <- emmeans(mod_7, specs = "exp", by = "isoform")

pairs(emmean7)

```

```{r model with isoform and replicate interaction }
#model (all replicates combined not separated by experiment)
mod_8 <- lm(log(pseudocounts) ~ factor(rep) * isoform, data = sum_filter1)
summary(mod_8)

#graph
sum_filter1 |>
  ggplot(aes(x = isoform, y = log(pseudocounts), color = factor(exp))) +
  geom_point(color = "lightgray")+
  stat_summary()+
  coord_flip()+
  labs(title = "Differences between experiments of the top 10 most abundant transcripts",
       x = "Isoform", y = "log(Pseudocount)", col = "Experiment")

# ggsave("/Users/danielle.dyer/Desktop/expdiff600.png", dpi = 600)

emmean8 <- emmeans(mod_8, specs = "rep", by = "isoform")

pairs(emmean8)


```


Looking at the difference between replicates
```{r model only including data from experiment 1}
#mod_9 isoform and replicate interaction with experiment 1 data
sum_filter1_exp1 <- sum_filter1|>
  filter(exp == "1")

mod_9 <- lm(log(pseudocounts) ~ factor(rep) * isoform, data = sum_filter1_exp1)
summary(mod_9)

emmean9 <- emmeans(mod_9, specs = "rep", by = "isoform")

pairs(emmean9)
```

```{r models only including data from experiment 2}
#mod_10 isoform and replicate interaction with experiment 2 data
sum_filter1_exp2 <- sum_filter1|>
  filter(exp == "2")

mod_10 <- lm(log(pseudocounts) ~ factor(rep) * isoform, data = sum_filter1_exp2)
summary(mod_10)

emmean10 <- emmeans(mod_10, specs = "rep", by = "isoform")

pairs(emmean10)
```

```{r graph difference between replicates facited by experiment}
sum_filter1 |>
  ggplot(aes(x = isoform, y = log(pseudocounts), color = factor(rep))) +
  geom_point(color = "lightgray")+
  stat_summary()+
  facet_wrap(vars(exp))+
  coord_flip()+
  labs(title = "Differences between replicates of the top 10 most abundant transcripts",
       x = "Isoform", y = "log(Pseudocount)", col = "Replicate")

```



Making animation to showing pseudocount change over days for presentation
```{r making an animation of change in psuedocounts of top 31 transcripts over the 11 days}
library(tidyverse)
library(gganimate)
# install.packages(gifski)
library(gifski)

MES1|>
  group_by(isoform) |>
  summarize(mean = mean(pseudocounts, na.rm = T)) |>
  filter(mean > 1000) ->
  ind_filter

MES1|>
  semi_join(y = ind_filter) ->
  mes_ind_filter

animation <- mes_ind_filter |>
  ggplot() +
  geom_tile(aes(x = rep, y = isoform, fill = log(pseudocounts))) +
  facet_wrap(~ exp) +
  scale_x_continuous(breaks = 1:2) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  transition_time(time = day) +
  labs(title = "Day {frame_time}", y = NULL)

animation <- animate(animation, renderer = gifski_renderer())

animation
```
